---
title: "Poisson log-normal model for fish species: network inference"
author: S. Robin
date: nov 2021
output: html_document
# output: pdf_document
# geometry: margin=1.5cm
# classoption: a4paper
---

## Data

* $n =89$ stations (sites) in the Barents' sea
* $m = 30$ fish species
* $d = 4$ environmental covariates: latitude, longitude, depth, temperature

```{r data, echo=TRUE}
rm(list=ls()); palette('R3')
data <- read.table("BarentsFish.csv", header=TRUE, sep=';')
dataName <- 'BarentsFish'
rownames(data) <- data[, 1]; data <- data[, -1]
X <- as.matrix(data[, 1:4]); Y <- as.matrix(data[, 5:ncol(data)])
n <- nrow(Y); m <- ncol(Y); d <- ncol(X)
print(head(X))
print(head(Y[, 1:10]))
```

```{r plotFun, echo=TRUE, message=FALSE}
library(fields)
PlotCov <- function(fit, cluster=NULL){
  if(is.null(cluster)){order <- 1:nrow(fit$model_par$Sigma)}else{order <- order(cluster)}
  Sigma <- as.matrix(fit$model_par$Sigma)[order, order]; Sigma[which(Sigma==0)] <- NA
  Omega <- as.matrix(fit$model_par$Omega)[order, order]; Omega[which(Omega==0)] <- NA
  par(mfrow=c(1, 2))
  image.plot(1:m, 1:m, cov2cor(Sigma), xlab='species', ylab='species', main='correlations')
  abline(h=.5+cumsum(table(cluster)), v=.5+cumsum(table(cluster)))
  image.plot(1:m, 1:m, cov2cor(Omega), xlab='species', ylab='species', main='conditional correlations')
  abline(h=.5+cumsum(table(cluster)), v=.5+cumsum(table(cluster)))
}
library(sna)
PlotNet <- function(fit, color=rep(1, nrow(fit$model_par$Omega))){
  adj <- 1*(fit$model_par$Omega!=0)
  gplot(adj, gmode='graph', vertex.col=color, vertex.border=color, edge.lwd=abs(fit$model_par$Omega), edge.col=8)
}
```

## PLN model

### Diagonal covariance matrix

No interaction: $\Omega$ and $\Sigma$ diagonal

```{r plnDiag, echo=TRUE, message=FALSE}
library(PLNmodels)
plnDiag <- PLN(Y ~ X, control=PLN_param(trace=0, covariance='diagonal'))
plnDiagCrit <- c(unlist(plnDiag$criteria), c(n_edges=0, EBIC=plnDiag$criteria$BIC, pen_loglik=plnDiag$criteria$loglik, density=0))
print(plnDiagCrit)
PlotCov(plnDiag)
```

### Full covariance matrix

All interactions: $\Omega$ and $\Sigma$ full

```{r plnFull, echo=TRUE, message=FALSE}
plnFull <- PLN(Y ~ X, control=PLN_param(trace=0, covariance='full'))
plnFullCrit <- c(unlist(plnFull$criteria), c(n_edges=m*(m-1)/2, EBIC=plnFull$criteria$BIC, pen_loglik=plnFull$criteria$loglik, density=1))
print(plnFullCrit)
PlotCov(plnFull)
```

### Abiotic effects 

```{r abiotic, echo=TRUE, message=FALSE}
Xsd <- apply(X, 2, sd)
beta <- diag(Xsd) %*% (plnFull$model_par$B[-1, ])
cluster <- kmeans(t(beta), centers=5, iter.max=1e2, nstart=1e2)$cluster
par(mfrow=c(1, 3))
image.plot(1:d, 1:m, plnFull$model_par$B[-1, ], main='abiotic effects', xlab='covariates', ylab='species')
text(1:d, rep(2, d), labels=substr(colnames(X), 1, 4), col=0)
image.plot(1:d, 1:m, beta, main='normalized effects', xlab='covariates', ylab='species')
text(1:d, rep(2, d), labels=substr(colnames(X), 1, 4), col=0)
image.plot(1:d, 1:m, beta[, order(cluster)], main='reordering species', xlab='covariates', ylab='species')
abline(h=.5+cumsum(table(cluster)), col=0, lwd=2)
text(1:d, rep(2, d), labels=substr(colnames(X), 1, 4), col=0)
```

## Network inference

Few interactions: $\Omega$ sparse

```{r net0, echo=TRUE, message=FALSE}
plnNet <- PLNnetwork(Y ~ X, control=PLNnetwork_param(trace=0))
plot(plnNet)
```

Extending the grid for the penalty parameter $\lambda$

```{r net1, echo=TRUE, message=FALSE}
lambda <- max(plnNet$penalties)^(seq(-5, 1, length.out=50))
plnNet <- PLNnetwork(Y ~ X, penalties=lambda, control=PLNnetwork_param(trace=0))
plot(plnNet)
```

### A first network

```{r netBic, echo=TRUE, message=FALSE}
netBic <- plnNet$getBestModel(crit='BIC')
print(netBic$criteria)
PlotCov(netBic)
par(mfrow=c(1, 2))
plot(netBic)
PlotNet(netBic, color=cluster)
```

### Stability selection

Resampling procedure to assess edge robustness

```{r netStab, echo=TRUE, message=FALSE}
stability_selection(plnNet)
netStab <- plnNet$getBestModel(crit='StARS')
print(netStab$criteria)
par(mfrow=c(1, 2))
plot(netStab)
PlotNet(netStab, color=cluster)
```

## Model comparison

```{r models, echo=TRUE, message=FALSE}
print(rbind(plnDiag = plnDiagCrit, 
            plnFull = plnFullCrit,
            netBic = netBic$criteria, 
            netStab = netStab$criteria))
```

Comparing estimated abiotic effects

```{r abiotics, echo=TRUE, message=FALSE}
abiotics = cbind(plnDiag = as.vector(diag(Xsd) %*% plnDiag$model_par$B[-1, ]), 
            plnFull = as.vector(diag(Xsd) %*% plnFull$model_par$B[-1, ]),
            netBic = as.vector(diag(Xsd) %*% netBic$model_par$B[-1, ]), 
            netStab = as.vector(diag(Xsd) %*% netStab$model_par$B[-1, ]))
plot(as.data.frame(abiotics), pch=20)
```

## Not accounting for environment

### PLN model

```{r plnNoCovar, echo=TRUE, message=FALSE}
plnNoCovar <- PLN(Y ~ 1, control=PLN_param(trace=0))
PlotCov(plnNoCovar)
PlotCov(plnNoCovar, cluster=cluster)
plnNoCovarCrit <- c(unlist(plnNoCovar$criteria), 
                    c(n_edges=m*(m-1)/2, EBIC=plnNoCovar$criteria$BIC,
                      pen_loglik=plnNoCovar$criteria$loglik, density=0))
```

### Network inference

```{r netNoCovar, echo=TRUE, message=FALSE}
plnNetNoCovar <- PLNnetwork(Y ~ 1, penalties=lambda, control=PLNnetwork_param(trace=0))
netNoCovarBic <- plnNetNoCovar$getBestModel(crit='BIC')
PlotCov(netNoCovarBic, cluster=cluster)
stability_selection(plnNetNoCovar)
netNoCovarStab <- plnNetNoCovar$getBestModel(crit='StARS')
PlotCov(netNoCovarStab, cluster=cluster)
```
